---
title: "Migration Experiment_01-Data-Exploration"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

#  I. Setup
## I.A. Libraries and Load Data
```{r Setup, include=FALSE}
library(tidyverse)
library(tibble)
library(here)
library(lubridate)
library(wesanderson)
library(viridisLite)
library(ggridges)
library(patchwork)
library(purrr)

# install.packages("tigris")  # one-time
library(auk)
library(ebirdst)  #ebirdst_runs contains breeding data
library(tigris)
library(terra)

# Load Data cleaned and processed by 00_data-cleaning
myebird_EXP <- readRDS("./data/derived/myebird_EXP.rds")
myebird_EXP_extra_obs_raw <- readRDS("./data/derived/myebird_EXP_extra_obs.rds") 
myebird_EXP_extra_obs <- myebird_EXP_extra_obs_raw %>% mutate(Count = as.numeric(Count)) # for some reason count is importing as a character

checklists_EXP <- readRDS("./data/derived/checklists_EXP.rds")
my_tax_EXP <- readRDS("./data/derived/my_taxonomy_EXP.rds")

# Birdcast
birdcast <- read.csv("./data/raw/Birdcast summary.csv")

```

## I.B. Useful Reference Values
```{r}
# Metadata
num_checklists = n_distinct(checklists_EXP$Checklist.ID)
num_pairs = n_distinct(checklists_EXP$Date)
obs_dates = unique(checklists_EXP$Day)


# List of Species
species_order <- my_tax_EXP %>% arrange(Taxonomy) %>% pull(Species)
num_species = length(species_order)
```


## I.D. Add Population column

"Population" is the max count for each particular day (defined in my filter function)

Adding to overall myebird_EXP dataframe, but if the filter function is used, it will be recalculated within the function

```{r}
myebird_EXP <- myebird_EXP %>% group_by(Date, Species) %>%
    mutate(Population = max(Count), .after = Count) %>% ungroup()
```


## I.E. Add Species Groups to myebird_EXP; Make Species and Groups factors

```{r}
# Add Species Groups to myebird_EXP
my_tax_EXP_join <- my_tax_EXP %>% select(Species, Species.Group)

myebird_EXP <- left_join(myebird_EXP, my_tax_EXP_join, by = "Species") %>%
  relocate(Species.Group, .after = Species)

# make Species and Groups factors
#... for myebird_EXP
myebird_EXP <- myebird_EXP %>% mutate(
  Species = factor(Species, levels = my_tax_EXP$Species),
  Species.Group = factor(Species.Group, levels = unique(my_tax_EXP$Species.Group)))
#... and for mytaxonomy
my_tax_EXP <- my_tax_EXP %>% mutate(
  Species = factor(Species, levels = my_tax_EXP$Species),
  Species.Group = factor(Species.Group, levels = unique(my_tax_EXP$Species.Group)))

```


## I.F. Functions
```{r, FUNCTIONS}
# ---- # MY FILTER FUNCTION ---- 
# - "keep": can select character vectors and/or specific species strings to keep 
#        - eg keep = c(vec1, "species1")
# - "group": associated with "Species.Group", which is a grouping of convenience for graphing, NOT a precise phylogenetic grouping 
# - "complete" and "paired" : pass TRUE/FALSE to function, if statement clears only if TRUE is passed, THEN filter is applied. 
#        - This shouldn't be necessary since dataframes are already segregated (myebird_EXP, myebird_EXP_extra_obs)
# - this function uses "my_tax_EXP" dataframe defined in setup chunk, outside of this function

my_filter <- function(df, 
                      complete = NULL, paired = NULL, 
                      keep = NULL, drop = NULL,  
                      group_keep = NULL, group_drop = NULL,
                      time_bin_keep = NULL,
                      date_drop = NULL) {
  
  
  # Fold group_keep into keep
  if(!is.null(group_keep)) {  
    group_sp <- my_tax_EXP %>% filter(Species.Group %in% group_keep) %>% pull(Species) %>% as.character()
    keep = c(keep, group_sp) }

  # Fold group_drop into drop
  if (!is.null(group_drop)) {
    group_sp <- my_tax_EXP %>% filter(Species.Group %in% group_drop) %>% pull(Species) %>% as.character()
    drop = c(drop, group_sp) }  
  
  
  # Save "df" argument as "result" so that it has a different name after all the filtering
  result <- df
  
  if(!is.null(complete))                result <- filter(result, Is.Complete == complete)
  if(!is.null(paired))                  result <- filter(result, Is.Paired == paired)
  if(!is.null(keep))          result <- filter(result, Species %in% keep)
  if(!is.null(drop))          result <- filter(result, !(Species %in% drop))
  # don't actually need the parentheses for !(Species %in% drop), bc !Species doesn't make sense, but !Boolean does
  if(!is.null(time_bin_keep)) result <- filter(result, Start.Time.Bin %in% time_bin_keep)
  if(!is.null(date_drop))     result <- filter(result, !(Date %in% date_drop))
  
 # result <- result %>% group_by(Date, Species) %>%
  #  mutate(Population = max(Count), .after = Count) %>% ungroup()
  
  result
}
```


```{r}
# ---- 0. REGION POLYGON: needed as the default `region_poly` for
# get_expected_abundance() below. Suffolk County, MA is loaded once here,
# then reprojected to match the CRS used by eBird's abundance rasters
# (we load one raster, "yerwar", purely to read its CRS off of it).
suffolk <- counties(state = "MA", cb = TRUE, progress_bar = FALSE) %>%
  filter(NAME == "Suffolk")

abd <- load_raster("yerwar", product = "abundance", resolution = "27km")

suffolk_v <- vect(suffolk)
suffolk_v <- project(suffolk_v, crs(abd))   # match raster's CRS


# DOWNLOAD SPECIES ABUNDANCE DATA AND MAKE 3-PART GRAPH

# ---- 1. FUNCTION DEFINITION: one species -> expected abundance curve ----
# manual overrides for species whose codes don't match my taxonomy df
code_overrides <- c(
  "Eastern Warbling Vireo" = "warvir",   # <- replace with what the search returns
  "Northern House Wren"    = "houwre"    # <- replace with what the search returns
)
get_expected_abundance <- function(species_common_name,
                                   taxonomy_df = my_tax_EXP,
                                   region_poly = suffolk_v,
                                   months_keep = c(3,4,5,6)) {
  code <- if (species_common_name %in% names(code_overrides)) {
    code_overrides[[species_common_name]]          # override wins if present
  } else {
    taxonomy_df %>% filter(Species == species_common_name) %>% pull(Species.Code)
  }
  # FIXED GUARD: catches NA, empty, and multi-match
  if (length(code) != 1 || is.na(code) || code == "") {
    warning("No usable species code for: ", species_common_name); return(NULL)
  }
  result <- try({
    ebirdst_download_status(species = code)
    load_raster(code, product = "abundance", resolution = "27km")
  }, silent = TRUE)
  if (inherits(result, "try-error")) {             # keeps one bad species from killing the whole run
    warning("Download/load failed for: ", species_common_name, " (", code, ")")
    return(NULL)
  }
  vals <- as.numeric(extract(result, region_poly, fun = mean, na.rm = TRUE)[1, -1])
  data.frame(Species = species_common_name,
             date = as.Date(names(result)),
             expected_abundance = vals) %>%
    filter(month(date) %in% months_keep) %>%
    mutate(yday = yday(date))
}


# ---- 2. FUNCTION DEFINITION: SHARED PLOT SETUP ----

# now a FUNCTION, since the x-range varies by species
align_layers <- function(rng) {
  list(
    geom_vline(xintercept = obs_yday, color = "grey65", linewidth = 0.3, linetype = "dotted"),
    annotate("rect", xmin = -Inf, xmax = min(obs_yday), ymin = -Inf, ymax = Inf,
             fill = "grey85", alpha = 0.4),
    annotate("rect", xmin = max(obs_yday), xmax = Inf, ymin = -Inf, ymax = Inf,
             fill = "grey85", alpha = 0.4),
    scale_x_continuous(breaks = month_breaks, labels = month_labels),
    coord_cartesian(xlim = rng)          # crops the view without dropping rows (no warnings)
  )
}

# ---- 3. FUNCTION DEFINITION: one species -> stacked 3-panel figure ----
make_3panel <- function(sp) {
  exp_sp <- expected_all %>% filter(Species == sp)
  if (nrow(exp_sp) == 0) return(NULL)              # skip species that failed lookup

  rng <- if (!is.null(range_overrides[[sp]])) range_overrides[[sp]] else default_range

  p1 <- ggplot(exp_sp, aes(x = yday, y = expected_abundance)) +
    geom_line() + geom_point() +
    align_layers(rng) +
    labs(title = sp, subtitle = "Expected regional abundance (eBird 2023 modeled estimate)", x = NULL, y = "Expected") +
    theme_minimal()
    
    
  p2 <- birdcast %>%
    ggplot(aes(x = yday, y = Average.Birds.in.Flight)) +
    geom_line() + geom_point() +
    align_layers(rng) +
    labs(subtitle = "BirdCast migration", x = NULL, y = "Birds in flight") +
    theme_linedraw()

 
  p3 <- myebird_EXP %>% filter(Species == sp) %>%
    mutate(yday = yday(Date)) %>%
    ggplot(aes(x = yday, y = Population)) +
    geom_line(linetype = "dashed", color = "grey50", linewidth = 0.4) +
    geom_point(size = 2) +
    geom_rug(sides = "b", length = unit(0.04, "npc")) +
    align_layers(rng) +
    labs(subtitle = "Your observed counts (dashed = interpolated between visits)",
         x = NULL, y = "Count") +
    theme_minimal()

  p1 / p2 / p3
}

```


#   II. Subdividing Datasets (DAY POPULATION, NO TIMEPOINT)

Let's make some subgrouped dataframes with my filter function

Pass LONG FORM dataframes for graphing (heat maps at least)

## II. A. Migrants, Residents, Nonnative
```{r}
#### Tags ####
residents_sp <- my_tax_EXP %>% filter(Resident == TRUE) %>% pull(Species)
migrants_sp  <- my_tax_EXP %>% filter(Resident==FALSE) %>% pull(Species)
nonnative_sp <- c("European Starling","House Sparrow")

#### Datasets ####
# Selections will remove duplicate rows
residents    <- my_filter(df = myebird_EXP, keep = residents_sp)  %>% 
  select(-"Count") %>% distinct(Date, Species, .keep_all=TRUE)

migrants     <- my_filter(df = myebird_EXP, keep = migrants_sp)  %>% 
  select(-"Count") %>% distinct(Date, Species, .keep_all=TRUE)
```

## II. B. Warblers: Datasets and Tags
```{r}
warblers    <- my_filter(df = myebird_EXP, group_keep = c("Warblers"))  %>% 
    select(-"Count") %>% distinct(Date, Species, .keep_all=TRUE)

warblers_sp <- warblers %>% distinct(Species) %>% pull(Species) %>% as.character()

warblers_uncommon_sp <- c("Common Yellowthroat", "Cape May Warbler","Magnolia Warbler", "Bay-breasted Warbler", "Blackburnian Warbler", "Northern Yellow Warbler", "Chestnut-sided Warbler", "Canada Warbler", "Black-throated Blue Warbler")

warblers_common_sp <- my_filter(warblers, drop = warblers_uncommon_sp) %>% distinct(Species) %>% pull(Species)

warblers_common <- my_filter(df= warblers, keep =  warblers_common_sp) 

######### REVIEW THIS CONCEPTUALLY (draw it out) BEFORE DELETING
# Table version of heat map! 
# Species column 1 (species = first rows) -> Each Date its own column (Widened by Date)
# Values under each date == population
# Passing only one timepoint bc population is duplicated
warblers_heatmap_to_table <- warblers %>% 
  select(Date, Population, Species) %>%
  pivot_wider(names_from = Date, values_from = Population)
#write_csv(warblers_heatmap_to_table, here("data", "warblers_heatmap_to_table.csv"))

```



##  II. D. Favorite Species - Datasets and Tags
### II.D.1. Favorite Species - Large List

- starting with all species, not just migrants

```{r}
favorites <- my_filter(myebird_EXP, drop = c("Mourning Dove", "Turkey Vulture", "Northern Mockingbird", "American Goldfinch", "Brown-headed Cowbird", "Common Grackle", "Northern Cardinal", "European Starling", "American Robin"), group_drop = c("Cormorants", "Corvids"))
favorites_sp <- favorites %>% distinct(Species) %>% pull(Species) %>% as.character()

# favorites_sp narrowed to migratory (non-resident) species only. Computed
# here (instead of down in the wave-delta section where it's also used) so
# the richness calculation below can use it too.
favorites_nonresident_sp <- favorites_sp[favorites_sp %in% as.character(migrants_sp)]
```

### II.D.3. Favorite Species - w Good Patterns

Specifically looking for clear patterns, multiple sightings, excluding boring birds (robins, mockingbirds)

```{r}
##### Dataset ###

good_patterns <- my_filter(myebird_EXP,
                           drop = c(residents_sp, nonnative_sp, warblers_uncommon_sp,"Swainson's Thrush","Blue-headed Vireo","Lincoln's Sparrow", "Rose-breasted Grosbeak", "Northern Mockingbird", "American Robin","Tree Swallow", "Blue-gray Gnatcatcher", "Brown-headed Cowbird", "Red-winged Blackbird", "Common Grackle", "Eastern Bluebird", "Gray Catbird", "Eastern Phoebe", "Baltimore Oriole", "Carolina Wren", "Northern Cardinal"), 
                           group_drop = c("Chickadees and Titmice","Waterfowl", "Grouse, Quail, and Allies", "Pigeons and Doves", "Cormorants", "Herons", "Corvids", "Hawks and Vultures", "Woodpeckers", "Finches", "Nuthatches", "Treecreepers", "Hummingbirds","Sparrows", "Waxwings" )) %>%
  select(-"Count") %>% distinct(Date, Species, .keep_all=TRUE)

# Organizing Species order
good_patterns <- good_patterns %>%
  mutate(Species = factor(Species, levels = c(
    "Great Crested Flycatcher",
    "Eastern Warbling Vireo",
    "Northern House Wren",
    "Wood Thrush",
    "Cedar Waxwing",
    "Eastern Wood-Pewee",
    "Scarlet Tanager",
    "Red-eyed Vireo",
    "Indigo Bunting",
    "Chimney Swift",
    "Pine Warbler",
    "Black-and-white Warbler",
    "Ovenbird",
    "Northern Parula",
    "Yellow-rumped Warbler",
    "Black-throated Green Warbler",
    "Blackpoll Warbler",
    "American Redstart"
  )))
```

#   III. Calculate Species Richness

## III.A. Make Richness Dataframes (wide version, each count gets its own column)

MAIN PURPOSE: Date -> # Unique Species

One row per day, so no time bins, species-specific counts/population

Will make different richness counts, based on different subsets of birds

- "All" richness       = Tot.Sp
- "Favorites" richness = Tot.Sp.Faves = only counting from my "favorite" species list
- "Non-Resident Favorites" richness = Tot.Sp.Nonres.Faves
- "Warblers" richness  = Tot.Sp.Warblers

```{r}
# 1. Start with richness Tot.Sp count for ALL birds
richness_by_day <- myebird_EXP %>% filter(Population>0) %>% group_by(Date) %>% 
  mutate(Tot.Sp = n_distinct(Species)) %>% ungroup() %>%
  select(c(Date:Day, Tot.Sp)) %>%
  unique()

# 2. get counts for other subsets, then join to richness_by_day df

# FAVORITES
richness_faves <- myebird_EXP %>% filter(Population>0, Species %in% favorites_sp) %>% group_by(Date) %>% 
  mutate(Tot.Sp.Faves = n_distinct(Species)) %>% ungroup() %>%
  select(c(Date, Tot.Sp.Faves)) %>%
  unique()

# MIGRANTS
richness_favorites_nonresident <- myebird_EXP %>% filter(Population>0, Species %in% favorites_nonresident_sp) %>% group_by(Date) %>%
  mutate(Tot.Sp.Nonres.Faves = n_distinct(Species)) %>% ungroup() %>%
  select(c(Date, Tot.Sp.Nonres.Faves)) %>%
  unique()

# WARBLERS
richness_warblers <- myebird_EXP %>% filter(Population>0, Species %in% warblers_sp) %>% group_by(Date) %>% 
  mutate(Tot.Sp.Warblers = n_distinct(Species)) %>% ungroup() %>%
  select(c(Date, Tot.Sp.Warblers)) %>%
  unique()


# 3. Joins!!

richness_by_day <- left_join(richness_by_day, richness_faves, by = "Date")
richness_by_day  <- left_join(richness_by_day, richness_favorites_nonresident, by = "Date")
richness_by_day <- left_join(richness_by_day, richness_warblers, by = "Date")

```

## III.B. LENGTHEN Richness df for faceted graphing! 

```{r}
richness_long <- richness_by_day %>% 
  pivot_longer(
    cols = c(Tot.Sp, Tot.Sp.Faves, Tot.Sp.Nonres.Faves, Tot.Sp.Warblers),
    names_to = "Species.Subset",
    values_to = "Total.Species"
    )
```

#   IV. Graphing
##  IV. A. GGPLOT THEMES and Other Prep

```{r, Themes}
# Heatmap for overall population by day-- no faceting by timepoint!
# for larger species counts (duplicated x axis)
heatmap_theme <- list(
  geom_tile(),
  theme_light(),
  scale_y_discrete(limits = rev),
  labs(
    x = NULL,
    y = NULL,
    fill = "Population"),
  theme( plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5)))

myviridis3 <-list(
  scale_fill_gradientn(
    colours = c(viridis(5)),
    values = scales::rescale(c(0, 3, 6, 9, 12, 24)))
)

myviridis_100 <-list(
  scale_fill_gradientn(
    colours = c(viridis(5)),
    values = scales::rescale(c(0, 5, 10, 20, 40, 80)))
)

```


##  IV. B. Color schemes

```{r}
warbler_colors <- c(
  "Ovenbird" = "#8C510A",                     # warm brown
  "Black-and-white Warbler" = "#000000",       # black
  "American Redstart" = "#E66101",             # burnt orange
  "Northern Parula" = "#56B4E9",                # sky blue
  "Blackpoll Warbler" = "#A69488",              # warm gray, orange-tinted (legs!)
  "Pine Warbler" = "#A6B32E",                   # olive-yellow-green
  "Yellow-rumped Warbler" = "#F0E442",          # bright lemon yellow
  "Black-throated Green Warbler" = "#1B7837",    # forest green
# -- new --
  "Black-throated Blue Warbler" = "#2166AC",   # deep steel blue (male plumage)
  "Northern Yellow Warbler" = "#E8A020",        # warm golden yellow (distinct from lemon #F0E442)
  "Blackburnian Warbler" = "#FF4500",          # vivid orange-red (iconic throat; brighter/redder than Redstart)
  "Bay-breasted Warbler" = "#8B1A1A",          # dark chestnut red (the bay flanks)
  "Magnolia Warbler" = "#C4C000",              # yellow-olive (yellow underparts)
  "Cape May Warbler" = "#D6604D",              # terracotta (rufous cheek)
  "Common Yellowthroat" = "#D4B200",           # warm yellow (yellow throat)
"Chestnut-sided Warbler" = "#C1692A"         # russet chestnut (the sides)
)
 
  


```


##  IV.C. Experiment Dates

```{r}
# Graph representing the experiment dates
myebird_EXP %>%
ggplot(aes(x=Date, y = Start.Time.Bin)) +
  geom_point(size = 4) +
  theme(axis.text.x = element_text(angle=45, size = 10, hjust=1),
          panel.grid.minor = element_blank()) +
  scale_x_date(breaks = sort(unique(myebird_EXP$Date)),
               date_labels = "%m-%d")+
  labs(title = "Distribution of Paired, Complete Checklists Across Dates")

 ggsave(file.path(fig_dir, sprintf("%s_day%02d.png", gsub(" ", "", wave_name), d)),
             p, width = 11, height = 7, dpi = 150)

```

##  IV. D. Birdcast Migration Data
### IV. D.1 Migration Data: Make lists, dfs, and lines
```{r}
# ---- WAVE BOUNDS: single source of truth for the 5 migration waves ----
# Every band/rectangle drawn later (on the "Day of May" x-axis used by
# BirdCast, and on the "day of year" x-axis used by everything else) is
# derived from this one table, so a wave's date range only ever needs to
# be edited in one place. (assign_wave(), further down in section IV.G,
# encodes the same 5 waves for per-day lookup and should be kept in sync
# if these ranges change.)
wave_bounds <- tribble(
  ~Wave,     ~xmin, ~xmax,
  "Wave 1",   4.6,   6.4,
  "Wave 2",  10.6,  11.4,
  "Wave 3",  12.6,  13.4,
  "Wave 4",  16.6,  19.4,
  "Wave 5",  25.6,  27.4
)

# Grey rectangles marking each wave, on the "Day of May" axis
band_peaks <- pmap(wave_bounds, function(Wave, xmin, xmax) {
  annotate("rect", xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf,
           fill = "grey50", alpha = 0.2)
})

# Same bands, shifted onto "day of year" (yday) coordinates -- used by every
# plot below that isn't on BirdCast's own "Day of May" axis.
may_offset <- yday(as.Date("2026-05-01")) - 1
wave_bands <- pmap(wave_bounds, function(Wave, xmin, xmax) {
  annotate("rect", xmin = xmin + may_offset, xmax = xmax + may_offset,
           ymin = -Inf, ymax = Inf, fill = "grey50", alpha = 0.2)
})

# Add Column Documenting Wave #
birdcast <- birdcast %>% mutate(
  Wave = factor(
    case_when(
#      Day.of.May == 1 ~ "Wave 1",
      Day.of.May == 5 | Day.of.May == 6 ~ "Wave 1",
      Day.of.May == 11 ~ "Wave 2",
      Day.of.May == 13 ~ "Wave 3",
      Day.of.May == 17 | Day.of.May == 18 | Day.of.May == 19 ~ "Wave 4",
      Day.of.May == 26 | Day.of.May == 27 ~ "Wave 5",
      .default = "No Wave"),
    levels = c("Wave 1", "Wave 2", "Wave 3", "Wave 4", "Wave 5", "Wave 6", "No Wave"),
    ordered = TRUE)
)
birdcast <- birdcast %>%
  mutate(yday = Day.of.May + (yday(as.Date("2026-05-01")) - 1))

```

### IV. D.2 Migration Data: Graphing

```{r}
title_size <- 18
subtitle_size <- 12
axis_title_size <- 12
axis_text_size <- 10
# Line graph Avg Migration
p_birdcast <- 
birdcast %>% filter(Day.of.May >=3 & Day.of.May <=29) %>%
                ggplot(aes(x=Day.of.May, y = Average.Birds.in.Flight))+
                geom_point() +
                geom_line()+
                band_peaks +
                theme_linedraw() +
                labs(title = "Suffolk County Birdcast Data",
                                    subtitle = "(Wave Peaks Highlighted in Grey)",
                                    x = "Morning After Overnight Migration (Day of May)",
                                    y = "Average Birds in Flight") +
                scale_x_continuous(breaks = seq(3,29,1), 
                                    expand = expansion(mult = c(0.03,0.03)))+
# Scale for patchworking
                scale_y_continuous(limits = c(0,15000),
                                   breaks = seq(0,15000,5000),
                                   expand = expansion(mult= c(0,0)))+
                theme(plot.title = element_text(hjust = 0.5, size = title_size),
                      plot.subtitle = element_text(hjust = 0.5, size = subtitle_size),
                      
                      axis.title.x=element_text(size = axis_title_size),
                      axis.text.x = element_text(size = axis_text_size, margin=margin(b=6, t=4), hjust = 0.5),
                      
                      axis.title.y = element_text(size =axis_title_size),
                      axis.text.y = element_text(size = axis_text_size, margin = margin(l=5),hjust = 0.5),
                      panel.grid.minor = element_blank())

p_birdcast

```

##  IV. E. Observations: Heatmaps and Ridgeplots

###  IV.E.1.   All Species

A psuedolog transformation of the color scale is used, along with custom breaks, because a few species have very high counts

```{r, fig.height=12, fig.width = 7}
# ALL Observations in Heat map, divided into Migrants and Residents
myebird_EXP_grouped <- myebird_EXP %>%
  left_join(my_tax_EXP %>% select(Species, Resident), by = "Species") %>%
  mutate(Resident = factor(Resident, levels = c(FALSE, TRUE), labels = c("Migrant", "Resident")))

myebird_EXP_grouped %>%
  ggplot(aes(x = Month.Day, y = Species, fill = Population)) +
  heatmap_theme +
   labs(title = "Day Population", 
       subtitle = "Migrants and Residents",
    #   x = "Date",
       y = NULL) +
  facet_grid(rows = vars(Resident), scales = "free_y", space = "free_y") +
  scale_fill_viridis_c(breaks = c(0,2,5,10,25,50,100), trans = "pseudo_log")+
 scale_x_discrete(sec.axis = dup_axis())+
  theme(axis.text.x.bottom = element_text(size=8, angle = 40, vjust = 0.6),
        axis.text.x.top = element_text(size=8, angle = 40, vjust = 0.4),
        axis.text.y = element_text(size = 8, vjust = 0.3))
```


###  IV.E.5.    Warblers: Heatmaps; Ridges
#### IV.E.5.i.  Warbler Heatmaps
```{r} 

# All Warblers
warblers %>%
ggplot(aes(x=Month.Day, y = Species, fill = Population))+
  heatmap_theme +
  labs(title = "Day Population", 
       subtitle = "Warblers!",
       x = "",
       y = NULL) +
  scale_fill_viridis_c(breaks = c(0,1, 2, 5,10,20),trans = "pseudo_log") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
  axis.text.x.top = element_text(margin = margin(b=4), hjust = 0))

  
```

#### IV.E.5.ii  Warbler Ridges

```{r, fig.height=8, fig.width=8}
# All warblers
p_warbs_ridges<- my_filter(warblers) %>%
                            ggplot(aes(x=Day, y = Species, height = Population, fill = Species))+
                            geom_ridgeline(scale = 0.2) +
                            scale_fill_manual(values = warbler_colors) +
                           # band_peaks +
                            theme_minimal() +
                            scale_x_continuous(breaks = obs_dates, 
                                               expand = expansion(mult = c(0.03,0.03)),
                                               limits = c(3,29)) +
                            labs(title = "\nWarbler Sightings",
                            subtitle = "(Colors correspond to actual bird colorings, \nwill change to be more readable later)",
                              x="Observation Day")+
                            theme(legend.position = "none",
                            plot.title = element_text(hjust = 0.5),
                            plot.subtitle = element_text(hjust = 0.5),
                                  panel.grid.major.x = element_line(color = "gray50", linewidth = 0.3),
                                  axis.ticks.minor.x = element_blank(),
                                  panel.grid.minor.x = element_blank(),
                                  axis.ticks.major.y = element_blank())
                            
p_birdcast / p_warbs_ridges + plot_layout(heights = c(1,6))

# Common warblers
p_warbs_common_ridges <- my_filter(warblers_common, drop = "Pine Warbler") %>%
                            ggplot(aes(x=Day, y = Species, height = Population, fill = Species))+
                            geom_ridgeline(scale = 0.2) +
                            scale_fill_manual(values = warbler_colors) +
                            band_peaks +
                            theme_minimal() +
                            scale_x_continuous(breaks = obs_dates, 
                                               expand = expansion(mult = c(0.03,0.03)),
                                               limits = c(3,29)) +
                            theme(legend.position = "none",
                                  panel.grid.major.x = element_line(color = "gray50", linewidth = 0.3),
                                  axis.ticks.minor.x = element_blank(),
                                  panel.grid.minor.x = element_blank(),
                                  axis.ticks.major.y = element_blank()) +
                            labs(title = "",
                              x="Observation Day")
#p_warbs_common_ridges

#p_birdcast / p_warbs_common_ridges + plot_layout(heights = c(1,6))

```

#### IV.E.5.iii Warbler DISCRETE Bar Graphs (technically more accurate than ridges)


```{r, fig.width=7, fig.height=5}
# STAGGERED BARS -- Use geom_rect so it can be merged like the ridgelines

# Give each species a numeric baseline position ####
warblers_common <- warblers_common %>%
    mutate(Species.Num = as.numeric(factor(Species)))

# GRAPHING
# Plot the Data ####
p_warbs_common_bars <- warblers_common %>%
    ggplot(aes(x = Day, y = Species.Num)) +
    geom_rect(aes(xmin = Day - 0.4, xmax = Day + 0.4,
                  ymin = Species.Num, ymax = Species.Num + Population * 0.15,
                  fill = Species)) +
# Theme and aesthetics ####
    theme_linedraw() +
    scale_fill_manual(values = warbler_colors)+
    band_peaks+
    labs(title = "",
         y = "")+
# Axes ####
    scale_x_continuous(
      limits = c(3,29),
      breaks = obs_dates,
      expand = expansion(mult= c(0.03,0.03))) +
    scale_y_continuous(
      breaks = sort(unique(warblers_common$Species.Num)),
      labels = levels(factor(warblers_common$Species)),
      expand = expansion(mult = c(0,0))) +
  # Use coord_cart instead of scale_y_cont(limits = ) so that geom_rect doesn't drop any bars if they go off the chart!
    coord_cartesian(ylim = c(1,10.7))+    
    theme(legend.position = "none",
         # panel.grid.major = seq(1,31,7),
          panel.grid.minor = element_blank())

p_birdcast / p_warbs_common_bars + plot_layout(heights= c(1,4))
```



###  IV.D.6.    18 Favorites: Ridges; Ridges+ Birdcast Merged PLot
#### IV.D.6.i.  Ridges plot alone

```{r}
# Make Ridges plot
p_good_patterns_ridges <- good_patterns %>%
                            ggplot(aes(x=Day, y = Species, height = Population, fill = Species))+
                            geom_ridgeline(scale = 0.15) +
                           # scale_fill_manual(values = warbler_colors) +
                           # band_peaks +
                            theme_minimal() +
                            scale_x_continuous(breaks = obs_dates, 
                                               expand = expansion(mult = c(0.03,0.03)),
                                               limits = c(3,29)) +
                            coord_cartesian(ylim = c(1.3, 17.4)) +
                            labs(title = "Favorite Birds, Clear Patterns",
                                 #subtitle = "",
                                 y = NULL,
                                 x="Observation Day") +                          
                            theme(legend.position = "none",
                                  plot.title = element_text(hjust = 0.5),
                                  panel.grid.major.x = element_line(color = "gray50", linewidth = 0.3),
                                  panel.grid.major.y = element_line(color = "gray0", linewidth = 0.3),
                               #  axis.ticks.major.y = element_blank()) +  # not defined in element hierarcy error
                                #  axis.ticks.minor.x = element_blank(),  # not defined in element hierarcy error
                                  panel.grid.minor.x = element_blank()) 

```

#### IV.D.6.ii.   Ridges + Birdcast Merged Plot 

```{r, fig.height=8, fig.width = 8}
# Re-set graph sizes for this merged figure specifically (smaller
# axis_title_size/axis_text_size than the standalone BirdCast plot above,
# since this version gets squeezed into a shared panel with the ridgeline).
title_size <- 18
subtitle_size <- 12
axis_title_size <-10
axis_text_size <-10

# Make Birdcast portion of plot
p_birdcast_large <-  p_birdcast + 
  theme(
    axis.title.y = element_text(size = axis_title_size, margin = margin(r = 0),vjust = 0.5),
    axis.text.y  = element_text(size = axis_text_size,  hjust = 1, margin = margin(r=3,l=0)),
    axis.title.x = element_text(size = axis_title_size)
                                )

# Make Ridges portion of plot
p_good_patterns_ridges_large <- 
  p_good_patterns_ridges +
  theme(plot.title = element_text(size=title_size, margin = margin(t=5)),
        axis.text.x = element_text(size=axis_text_size),
        axis.text.y = element_text(size=axis_text_size))

# Print plot
p_birdcast_large / p_good_patterns_ridges_large + plot_layout(heights = c(1,6))
```



##  IV. F. 3-panel figures


### IV. F.2. Download Expected Species Data (Abundance from)

```{r}
# ---- CALL MAP_DFR TO DOWNLOAD SPECIES DATA-- all 18 species ----
# Checks if it's already been downloaded
# Saves new RDS ONLY ONCE, expected_all and exp_path
good_patterns_sp <- levels(good_patterns$Species)

# Relative to this .Rmd's own folder (same convention as the readRDS() calls
# in Setup), so it still works no matter whose computer this project is on.
exp_path <- "./data/expected_all.rds"

if (file.exists(exp_path)) {
  expected_all <- readRDS(exp_path)
} else {
  expected_all <- map_dfr(good_patterns_sp, get_expected_abundance)
  saveRDS(expected_all, exp_path)
}

```


```{r}
#-------- 3-GRAPH/SPECIES; MANY SPECIES: BIRDCAST, EXPECTED=ABUNDANCE, ACTUAL OBSERVED  ---------

# ---- 1.  SHARED PLOT SETUP ----
obs_yday <- sort(unique(yday(myebird_EXP$Date)))        # days you actually birded

# default window: mid-April through mid-June
default_range <- yday(as.Date(c("2026-04-26", "2026-06-07")))

# species that need an earlier start (last ~2 weekly points of March)
range_overrides <- list(
  "Yellow-rumped Warbler" = yday(as.Date(c("2026-03-29", "2026-06-08"))),
  "Pine Warbler"          = yday(as.Date(c("2026-03-29", "2026-06-08")))
)

# per-species x windows: overrides where defined, default otherwise
# MOVED UP from the "2-GRAPHS" chunk further down -- trim_to_window() gets
# called by name earlier in this doc (the "1-GRAPH/SP" chunk below), so both
# rng_df and the function itself have to exist before that first call, not
# just before their original spot.
rng_df <- tibble(
  Species = factor(good_patterns_sp, levels = good_patterns_sp),
  lo = sapply(good_patterns_sp, function(s)
        if (!is.null(range_overrides[[s]])) range_overrides[[s]][1] else default_range[1]),
  hi = sapply(good_patterns_sp, function(s)
        if (!is.null(range_overrides[[s]])) range_overrides[[s]][2] else default_range[2])
)

# TRIM FIRST — so the denominator reflects only the visible window
trim_to_window <- function(df) {
  df %>%
    left_join(rng_df, by = "Species") %>%
    filter(yday >= lo, yday <= hi) %>%
    select(-lo, -hi)
}

month_breaks <- yday(as.Date(c("2026-03-01","2026-04-01","2026-05-01","2026-06-01")))
month_labels <- c("Mar","Apr","May","Jun")

# ---- 2. CALL 3-PANEL GRAPHING FUNCTION: one figure per species ----
all_figs <- map(good_patterns_sp, make_3panel)   # <- the call
names(all_figs) <- good_patterns_sp
all_figs[["Blackpoll Warbler"]]              # view one
all_figs[["Pine Warbler"]]                   # view one
#all_figs
```


```{r}
# CLAUDE

# ----- 1-GRAPH/SP; MANY SPECIES = ACTUAL OBSERVED COUNTS ONLY ----
# ---- MANY SPECIES PER FIGURE


month_breaks <- yday(as.Date(c(
  "2026-03-01","2026-03-15","2026-04-01","2026-04-15",
  "2026-05-01","2026-05-15","2026-06-01","2026-06-15")))
month_labels <- c("Mar 1","Mar 15","Apr 1","Apr 15","May 1","May 15","Jun 1","Jun 15")

# may_offset and wave_bands (the grey migration-wave rectangles in yday
# space) were already built once in section IV.D.1 -- reused here as-is.

obs_counts <- myebird_EXP %>%
  filter(Species %in% good_patterns_sp) %>%
  mutate(Species = factor(Species, levels = good_patterns_sp),
         yday = yday(Date)) %>%
  distinct(Species, yday, Population) %>%
  trim_to_window()

# same "only show extras that add information" logic, but on raw counts
interp_counts <- function(sp, x) {
  o <- obs_counts %>% filter(Species == sp) %>% arrange(yday)
  if (nrow(o) < 2) return(NA_real_)
  approx(o$yday, o$Population, xout = x, rule = 1)$y
}

extra_counts <- myebird_EXP_extra_obs %>%
  filter(Presence.Only == FALSE) %>%
  filter(Species %in% good_patterns_sp) %>%
  mutate(Species = factor(Species, levels = good_patterns_sp),
         yday = yday(Date)) %>%
  rowwise() %>%
  mutate(obs_interp = interp_counts(Species, yday)) %>%
  ungroup() %>%
  filter(is.na(obs_interp) | Count > obs_interp) %>%
  trim_to_window()

ggplot(obs_counts, aes(x = yday, y = Population)) +
  wave_bands +
  geom_line(linetype = "dashed", color = "grey50", linewidth = 0.4) +   # dashed = interpolated
  geom_point(color = "#1B7837", size = 1.4) +                           # solid = real observation
  geom_rug(sides = "b", length = unit(0.03, "npc"), color = "grey40") + # tick per sampling day
  geom_point(data = extra_counts,
             aes(x = yday, y = Count, shape = "Unpaired / incomplete checklist"),
             color = "red", size = 2, inherit.aes = FALSE) +
  scale_shape_manual(values = c("Unpaired / incomplete checklist" = 8)) +
  guides(shape = guide_legend(override.aes = list(color = "red"))) +
  facet_wrap(~Species, ncol = 3, scales = "free") +
  scale_x_continuous(breaks = month_breaks, labels = month_labels) +
  labs(title = "Observed Counts by Species",
       subtitle = "Actual counts, not normalized — y-axis differs per panel.\nGrey bands = BirdCast migration waves. Dashed line = interpolated between visits.",
       x = NULL, y = "Day population (max individuals seen)", shape = NULL) +
  theme_minimal() +
  theme(legend.position = "top",
        strip.text = element_text(size = 8, face = "bold"),
        strip.background = element_rect(fill = "grey90", color = NA),
        panel.spacing.y = unit(1, "lines"),
        panel.spacing.x = unit(0.5, "lines"))
```


```{r}
# CLAUDE

# ---- 2-GRAPHS = ACTUAL OBSERVED COUNTS &  EBIRD EXPECTED ABUNDANCE ----
# ----- MANY SPECIES PER FIGURE
# ---- SCALING METHOD: "max", "mean", or "total" ----

scale_method <- "mean"

y_lab <- switch(scale_method,
  max   = "Proportion of each series' own peak (0–1)",
  mean  = "Normalized to series average \n(1.0 = mean)",
  total = "Share of season total")

# may_offset and wave_bands (the grey migration-wave rectangles in yday
# space) were already built once in section IV.D.1 -- reused here as-is.

# rng_df and trim_to_window() are now defined earlier in the doc (1.SHARED
# PLOT SETUP chunk), since trim_to_window() was being called before it was
# defined. Left rng_df's construction there since good_patterns_sp,
# range_overrides, and default_range are already available by that point.

exp_long <- expected_all %>%
  transmute(Species, yday, value = expected_abundance, source = "Expected (regional)") %>%
  mutate(Species = factor(Species, levels = good_patterns_sp))

obs_long <- myebird_EXP %>%
  filter(Species %in% good_patterns_sp) %>%
  mutate(yday = yday(Date)) %>%
  distinct(Species, yday, Population) %>%
  transmute(Species, yday, value = Population, source = "Observed (mine)") %>%
  mutate(Species = factor(Species, levels = good_patterns_sp))

exp_long <- trim_to_window(exp_long)
obs_long <- trim_to_window(obs_long)

# THEN scale, using only what's displayed
combined <- bind_rows(exp_long, obs_long) %>%
  group_by(Species, source) %>%
  mutate(denom = case_when(
           scale_method == "max"   ~ max(value,  na.rm = TRUE),
           scale_method == "mean"  ~ mean(value, na.rm = TRUE),
           scale_method == "total" ~ sum(value,  na.rm = TRUE)),
         value_scaled = ifelse(denom > 0, value / denom, 0)) %>%
  ungroup() %>%
  select(-denom)

obs_denom <- obs_long %>%
  group_by(Species) %>%
  summarize(denom = case_when(
              scale_method == "max"   ~ max(value,  na.rm = TRUE),
              scale_method == "mean"  ~ mean(value, na.rm = TRUE),
              scale_method == "total" ~ sum(value,  na.rm = TRUE)),
            .groups = "drop")

interp_obs <- function(sp, x) {
  o <- combined %>%
    filter(source == "Observed (mine)", Species == sp) %>%
    arrange(yday)
  if (nrow(o) < 2) return(NA_real_)
  approx(o$yday, o$value_scaled, xout = x, rule = 1)$y
}

extra_scaled <- myebird_EXP_extra_obs %>%
  filter(Presence.Only == FALSE) %>%
  filter(Species %in% good_patterns_sp) %>%
  mutate(Species = factor(Species, levels = good_patterns_sp),
         yday = yday(Date)) %>%
  left_join(obs_denom, by = "Species") %>%
  mutate(value_scaled = ifelse(denom > 0, Count / denom, 0)) %>%
  rowwise() %>%
  mutate(obs_interp = interp_obs(Species, yday)) %>%
  ungroup() %>%
  filter(is.na(obs_interp) | value_scaled > obs_interp) %>%
  trim_to_window()

ggplot(combined, aes(x = yday, y = value_scaled, color = source, linetype = source)) +
  wave_bands +                                    # waves behind everything
  geom_line(linewidth = 0.5) +
  geom_point(size = 0.8) +
  geom_point(data = extra_scaled,
             aes(x = yday, y = value_scaled, shape = "Unpaired / incomplete checklist"),
             color = "red", size = 2, inherit.aes = FALSE) +

  # all three scales share name = NULL, which merges color + linetype into ONE legend
  scale_color_manual(name = NULL,
    values = c("Expected (regional)" = "grey40", "Observed (mine)" = "#1B7837")) +
  scale_linetype_manual(name = NULL,
    values = c("Expected (regional)" = "solid", "Observed (mine)" = "dashed")) +
  scale_shape_manual(name = NULL,
    values = c("Unpaired / incomplete checklist" = 8)) +
  # legend key would otherwise be black, since color is set outside aes()
  guides(shape = guide_legend(override.aes = list(color = "red"))) +

  facet_wrap(~Species, ncol = 3, scales = "free_y") +
  scale_x_continuous(breaks = month_breaks, labels = month_labels) +
  coord_cartesian(xlim = default_range) +

  labs(title = "Observed Population vs. Expected Abundance",
       subtitle = " 'Observed Population' = Counts from my outings\n'Expected Abundance' = from eBird Modeling\n  Grey bands = BirdCast migration waves",
       caption = "*Note*: 'Expected' is a multi-year average, so it's a smoothed estimate -- real migration comes in waves",
       x = NULL, y = y_lab) +

  theme_minimal() +
  theme(legend.position = "top",
        plot.title    = element_text(hjust = 0.5, size = 16, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 11),
        plot.caption  = element_text(hjust = 0.5, size = 11),
        strip.text = element_text(size = 8, face = "bold"),
        strip.background = element_rect(fill = "grey90", color = NA),  # boxes the label with its panel
        panel.spacing.y = unit(1, "lines"),                            # more gap between rows than within
        panel.spacing.x = unit(0.5, "lines"))
```



##  IV. G. Species Richness (Total species count)

```{r}
subset_labels <-c(
  "Tot.Sp"              =   "All Species",
  "Tot.Sp.Faves"        = "Favorite Species",
  "Tot.Sp.Nonres.Faves" = "Migrant Favorites",
  "Tot.Sp.Warblers"     = "Warblers"
)


richness_long %>% filter(!Species.Subset %in% c("Tot.Sp.Faves"))%>%
ggplot(aes(x=Month.Day, y = Total.Species, fill = Species.Subset))+
  geom_col() +
  facet_wrap(~Species.Subset, labeller = as_labeller(subset_labels)) +
  scale_fill_discrete(labels = subset_labels) +
  theme_linedraw() +
  theme(axis.text.x = element_text(size=6, angle = 45, vjust = 0.4)) +
  labs(title = "Species Richness by Day",
       y = "Number of Species",
       x = "Date")

richness_long %>% filter(!Species.Subset %in% c("Tot.Sp.Faves"))%>%
ggplot(aes(x=Month.Day, y = Total.Species, fill = Species.Subset))+
  geom_col(position = "dodge", color = "grey30") +
  scale_fill_discrete(labels = subset_labels)+
  theme_minimal() +
  labs(title = "Species Richness by Day",
       y = "Number of Species",
       x = "Date") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1))

```

```{r, fig.width = 11, fig.height = 7}
# ---- 1. WAVE-HIGHLIGHTABLE BAND PEAKS ----
# wave_bounds (one row per wave) was already defined in section IV.D.1 and
# is reused here as the single source of truth for every wave's bounds.

make_band_peaks <- function(highlight_wave = NULL) {
  pmap(wave_bounds, function(Wave, xmin, xmax) {
    if (!is.null(highlight_wave) && Wave == highlight_wave) {
      annotate("rect", xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf,
               fill = "#F5A623", alpha = 0.35, color = "#B8730A", linewidth = 1)
    } else {
      annotate("rect", xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf,
               fill = "grey50", alpha = 0.2)
    }
  })
}

# "Wave N" text sitting inside the top of each band, as in the slides
make_band_labels <- function(y = 13600, size = 2.7) {
  annotate("text",
           x = (wave_bounds$xmin + wave_bounds$xmax) / 2,
           y = y, label = wave_bounds$Wave,
           size = size, color = "grey20")
}

# ---- 2. BIRDCAST PLOT BUILDER ----
build_birdcast_plot <- function(bands = band_peaks, observed_days = NULL) {
  p <- birdcast %>% filter(Day.of.May >= 3 & Day.of.May <= 29) %>%
    ggplot(aes(x = Day.of.May, y = Average.Birds.in.Flight)) +
    bands +                     # bands behind the trace
    make_band_labels() +
    geom_point() +
    geom_line() +
    theme_linedraw() +
    labs(title = "Suffolk County Birdcast Data",
         subtitle = "(Wave Peaks Highlighted in Grey)",
         x = "Morning After Overnight Migration (Day of May)",
         y = "Average Birds in Flight") +
    scale_x_continuous(breaks = seq(3, 29, 1), expand = expansion(mult = c(0.03, 0.03))) +
    scale_y_continuous(limits = c(0, 15000), breaks = seq(0, 15000, 5000), expand = expansion(mult = c(0, 0))) +
    theme(plot.title = element_text(hjust = 0.5, size = title_size),
          plot.subtitle = element_text(hjust = 0.5, size = subtitle_size),
          axis.title.x = element_text(size = axis_title_size),
          axis.text.x = element_text(size = axis_text_size, margin = margin(b = 6, t = 4), hjust = 0.5),
          axis.title.y = element_text(size = axis_title_size),
          axis.text.y = element_text(size = axis_text_size, margin = margin(l = 5), hjust = 0.5),
          panel.grid.minor = element_blank())

  # optional day marker; the slide reference doesn't use one, so it's off by default
  if (!is.null(observed_days)) {
    p <- p + geom_vline(xintercept = observed_days, color = "black", linewidth = 0.7, linetype = "solid")
  }
  p
}

# ---- 3. WAVE ASSIGNMENT FOR YOUR OBSERVATION DAYS ----
assign_wave <- function(day) {
  case_when(
    day %in% c(5, 6)       ~ "Wave 1",
    day == 11              ~ "Wave 2",
    day == 13              ~ "Wave 3",
    day %in% c(17, 18, 19) ~ "Wave 4",
    day %in% c(26, 27)     ~ "Wave 5",
    TRUE                   ~ NA_character_
  )
}

# ---- 4. SPECIES x DAY GRID WITH DELTA ----
build_species_day_grid <- function(species_subset) {
  all_dates <- sort(unique(myebird_EXP$Date))
  myebird_EXP %>%
    select(Date, Species, Population) %>%
    distinct() %>%
    filter(Species %in% species_subset) %>%
    mutate(Species = as.character(Species)) %>%
    complete(Date = all_dates, Species = species_subset, fill = list(Population = 0)) %>%
    mutate(Species = factor(Species, levels = species_subset),
           Day = day(Date)) %>%
    arrange(Species, Date) %>%
    group_by(Species) %>%
    mutate(Delta = Population - lag(Population)) %>%
    ungroup()
}

# ---- 5. TAG GRID WITH WAVE ----
tag_wave_days <- function(grid_df) {
  grid_df %>%
    mutate(Wave = factor(assign_wave(Day), levels = wave_bounds$Wave)) %>%
    filter(!is.na(Wave)) %>%
    mutate(Wave = droplevels(Wave))
}

# ---- 6. WHICH OBSERVED DAYS FALL IN A GIVEN WAVE ----
obs_days_for_wave <- function(wave_name) {
  all_days <- sort(unique(myebird_EXP$Day))
  all_days[assign_wave(all_days) %in% wave_name]
}

# ---- 7. UNCOMMON SPECIES (< 4 total individuals across the whole experiment) ----
uncommon_sp <- myebird_EXP %>%
  select(Date, Species, Population) %>%
  distinct() %>%
  group_by(Species) %>%
  summarize(Total.Population = sum(Population), .groups = "drop") %>%
  filter(Total.Population < 4) %>%
  pull(Species) %>%
  as.character()

# ---- 8. FIXED AXIS FOR ALL WAVE-DELTA GRAPHS ----
# Same frame on every graph so waves stay comparable. Bars past the frame are
# CLAMPED to the frame edge (see y_draw below) and their label still prints the
# true count, so a handful of blown-out bars never force the axis to expand.
delta_ylim   <- c(-6, 11)                 # visible frame
delta_breaks <- c(-5, 0, 5, 10)           # axis ticks

# ---- 9. SPECIES SUBSET: FAVORITES MINUS RESIDENTS ----
# favorites_nonresident_sp was already computed in section II.D.1 (it's also
# used by the richness calculation above) -- reused here as-is.
grid_favorites_nonresident <- build_species_day_grid(favorites_nonresident_sp)
grid_favorites_nonresident_wave <- tag_wave_days(grid_favorites_nonresident)
warbler_sp_in_favorites_nonresident <- intersect(favorites_nonresident_sp, warblers_sp)

# ---- 10. ONE DAY'S BARS ----
plot_day_delta <- function(df_wave, wave_name, day, subset_label, warbler_sp = NULL) {
  df <- df_wave %>%
    filter(Wave == wave_name, Day == day, !is.na(Delta), Delta != 0) %>%
    mutate(Species = droplevels(Species),
           Is.Uncommon = Species %in% uncommon_sp)

  lab_off  <- 0.55   # in-range label sits this far past the bar tip
  cap_off  <- 1.30   # capped label sits this far INSIDE the frame edge,
                     # which also keeps it clear of the "Warblers" label below
  diam_off <- 1.7    # diamond sits this far past the LABEL, so they never collide

  df <- df %>%
    mutate(
      # y_draw is what geom_col actually draws.  clip = "off" is needed for the
      # "Warblers" bracket label, and it also stops ggplot from trimming the
      # bars, so an out-of-range bar has to be clamped in the DATA or it runs
      # straight off the bottom of the figure.
      y_draw  = pmin(pmax(Delta, delta_ylim[1]), delta_ylim[2]),
      # ">=" so a bar that lands exactly on the frame edge is labelled inside
      # the bar too, the way the slides do it
      Capped  = Delta >= delta_ylim[2] | Delta <= delta_ylim[1],
      # in-range bar: label just past the tip, in black.
      # capped bar: label tucked inside the bar in white, so it can never
      # collide with the title above or the species names below.
      label_y    = ifelse(Capped, y_draw - sign(Delta) * cap_off,
                                  y_draw + sign(Delta) * lab_off),
      label_col  = ifelse(Capped, "white", "black"),
      label_text = ifelse(Delta > 0, paste0("+", Delta), as.character(Delta)),
      diamond_y  = ifelse(Capped, y_draw - sign(Delta) * (cap_off + 1.1),
                                  y_draw + sign(Delta) * diam_off)
    )

  p <- ggplot(df, aes(x = Species, y = y_draw, fill = Delta > 0)) +
    geom_col(width = 0.65) +
    geom_hline(yintercept = 0, color = "grey40", linewidth = 0.3) +
    geom_text(aes(y = label_y, label = label_text, color = label_col),
              size = 2.9, fontface = "bold", show.legend = FALSE) +
    scale_color_identity() +
    scale_fill_manual(values = c(`TRUE` = "#2166AC", `FALSE` = "#B2182B"), guide = "none") +
    scale_y_continuous(breaks = delta_breaks, expand = expansion(mult = c(0, 0))) +
    coord_cartesian(ylim = delta_ylim, clip = "off") +
    labs(title = paste0(wave_name, ", Day ", day,
                        ": Change vs. Previous Observed Day — ", subset_label),
         x = NULL, y = "Change in Population") +
    theme_minimal() +
    theme(plot.title   = element_text(hjust = 0.5, face = "bold", size = 12,
                                      margin = margin(b = 14)),  # room for capped labels above the frame
          plot.margin  = margin(t = 8, r = 12, b = 8, l = 8),
          axis.text.x  = element_text(angle = 45, hjust = 1, size = 7.5,
                                      margin = margin(t = 4)),
          panel.grid.minor = element_blank())

  # diamond on every uncommon-species bar
  diamonds <- df %>% filter(Is.Uncommon)
  if (nrow(diamonds) > 0) {
    p <- p + geom_point(data = diamonds, aes(x = Species, y = diamond_y),
                        shape = 18, size = 3.4, color = "#F5A623", inherit.aes = FALSE)
  }

  # green bracket + label around the warbler block
  if (!is.null(warbler_sp) && length(warbler_sp) > 0) {
    warb_present <- intersect(levels(df$Species), warbler_sp)
    if (length(warb_present) > 0) {
      xs <- match(warb_present, levels(df$Species))
      x_min <- min(xs) - 0.5; x_max <- max(xs) + 0.5
      p <- p +
        annotate("rect", xmin = x_min, xmax = x_max,
                 ymin = delta_ylim[1], ymax = delta_ylim[2],
                 fill = NA, color = "#1B7837", linewidth = 0.7) +
        # label sits just INSIDE the frame, so it can't collide with the
        # species names below; the white fill keeps it readable if a bar
        # happens to run behind it
        annotate("label", x = (x_min + x_max) / 2, y = delta_ylim[1] + 0.42,
                 label = "Warblers", color = "#1B7837", fontface = "bold", size = 3.2,
                 fill = "white", linewidth = 0, label.padding = unit(0.12, "lines"))
    }
  }
  p
}

# ---- 11. BIRDCAST PANEL ABOVE THAT DAY'S DELTA BARS ----
plot_day_with_birdcast <- function(df_wave, wave_name, day, subset_label, warbler_sp = NULL) {
  p_top <- build_birdcast_plot(make_band_peaks(highlight_wave = wave_name)) +
    theme(plot.title = element_text(size = 13), plot.subtitle = element_text(size = 9))
  p_bottom <- plot_day_delta(df_wave, wave_name, day, subset_label, warbler_sp = warbler_sp)
  p_top / p_bottom + plot_layout(heights = c(1, 2.2))
}

# ---- 12. OUTPUT: ALL WAVES ----
# fig.width / fig.height live in THIS chunk header, so every wave gets the same
# frame. Keep the whole loop in one chunk; splitting it (or dropping the header
# options on one chunk) is what made the last wave render at a different size.
#
# RStudio's inline preview re-renders the most recently drawn plot at whatever
# size the pane happens to be, so the last graph in a loop can still LOOK
# smaller there. The files written below are the ones to trust, and they are
# what the portfolio page should use.
save_wave_figs <- TRUE
fig_dir        <- "figures/waves"
if (save_wave_figs) dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

for (wave_name in wave_bounds$Wave) {
  for (d in obs_days_for_wave(wave_name)) {
    p <- plot_day_with_birdcast(grid_favorites_nonresident_wave, wave_name, d,
                                "Non-Resident Favorites",
                                warbler_sp = warbler_sp_in_favorites_nonresident)
    print(p)
    if (save_wave_figs) {
      ggsave(file.path(fig_dir, sprintf("%s_day%02d.png", gsub(" ", "", wave_name), d)),
             p, width = 11, height = 7, dpi = 150)
    }
  }
}
```
